One-loop PDF of the average Ising magnetization in 4 dimensions.

### 2020-03-22 paper/paper.tex
Paper checkpoint

\title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
\author{Andrea Allais}
\emailAdd{allais.andrea@gmail.com}

\begin{document}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{f_functions.png}
\end{center}
\caption{\label{fig:f_functions} The functions $f_i$}
\end{figure}

@@

\section{Derivation}

@@

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
won't need any of its special properties beyond its asymptotic behavior, which
can be easily obtained from the definition.}
\begin{equation}
    \label{eq:theta}
    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}

@@

@@

\bibitem{WeinbergEffectiveAction}
S. Weinberg,
\textit{The quantum theory of fields, Vol. 2}, Cambridge University Press, (1996) pg. 63.
\end{thebibliography}

@@

At the critical point it is convenient to introduce the dimensionless 
quantity
\begin{equation}
    \bar{\varphi} =
    \left(\frac{2\pi^2}{9} V g\right)^{\frac{1}{4}} \bar{\phi}\,,
\end{equation}
whose probability distribution is:
\begin{equation}
\label{eq:log_pdf_critical}
\begin{split}
    -\log p(\bar{\varphi}) = \mathrm{const.} +  
    &g^{\frac{1}{2}}\bar{\varphi}^2 
      \left(\frac{\pi}{\sqrt{2}}\left(f_1(0) - 1\right)  
      + O\left(g\right) \right) + \\
    &\bar{\varphi}^4\left(
      1 - \frac{g}{2}f_2\left(0\right)
      + O\left(g^2\right)\right) + \\
    &g^{\frac{3}{2}} \bar{\varphi}^6 \left(\frac{f_3(0)}{3\sqrt{2} \pi}
    + O\left(g\right)\right) + O\left(g^2\bar{\varphi}^8\right)\,.
\end{split}
\end{equation}

This expression has a residual dependence on system size via dimensional
transmutation of the renormalized coupling $g$, which at the critical point $r
= 0$ follows the simplified renormalization group equation
\begin{equation}
    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,.
\end{equation}

Therefore, as system size grows at fixed bare couplings $r_0$ $u_0$, the
renormalized coupling $g$ goes to zero, and the all terms in
(\ref{eq:log_pdf_critical}) besides the quartic vanish. Thus we conclude
that, for sufficiently large system size: 
\begin{equation}
    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,.
\end{equation}

>>>>

\title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
\author[a]{Andrea Allais}
\affiliation[a]{Independent scholar}
\emailAdd{allais.andrea@gmail.com}
\abstract{We compute at one loop in perturbation theory the probability density
function of the average magnetization $\bar{\phi}$ of the Ising model on both
the 4-torus and the 4-sphere. The perturbative expansion covers the entire
symmetric phase and a neighborhood of the critical point. We find that, at the
critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
\exp(-f(R) \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
results by comparison with Monte Carlo simulation.}
\begin{document}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{f_functions.png}
\end{center}
\caption{\label{fig:f_functions} The functions $f_i$ that appear in
    (\ref{eq:log_pdf}). The functions go to zero as $x\to\infty$, and diverge
    for $x\to-1$ ($x\to-4$ for the sphere), signaling an instability of the
    perturbative vacuum.}
\end{figure}

@@

\section{Derivation on the 4-torus}

@@

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:
\begin{equation}
    \label{eq:theta}
    \theta(z) = \theta_3(0, e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}
Here $\theta_3$ is a Jacobi theta function, but we will not need any of
its special properties beyond its asymptotic behavior, which can be easily
obtained from the definition.

@@

\section{Results for the 4-sphere}
\label{sec:four_sphere}
It is possible to obtain $\log p(\bar{\phi})$ on the 4 sphere as well, with
very similar methods. Here we highlight the main differences from the 4 torus.

Curved manifolds allow for an additional renormalizable coupling: 
\begin{equation}
    S += \int \dd^4 x\ \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
\end{equation}
where $\mathcal{R}$ is the scalar curvature and $\xi_0$ is a dimensionless bare
coupling. On a static manifold, this coupling is equivalent to a mass term.
The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
presence of interactions, the coupling needs to be renormalized as well, and
its renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
renormalization group equation for $\xi$ is:
\begin{equation}
    \mu \frac{\dd \xi}{\dd \mu} =  \frac{1}{3} g \left(\xi - \frac{1}{6}\right)\,.
\end{equation}

From this experssion it seems that $\xi$ can be set to the critical value
$\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
at higher orders in perturbation theory the renormalization group equation
becomes inhomogeneous \cite{Brown1980}. Therefore, on a curved manifold, $\xi$
is simply another free parameter of the scalar field theory.

The probability distribution of the total magnetization on the 4-sphere is
still given by (\ref{eq:log_pdf}), with the substitution:
\begin{equation}
    rR^2 \to \frac{\xi}{12} + r R^2\,
\end{equation}
where $R$ is now the radius of the sphere, and with the functions $f_i$ defined
as:
\begin{align}
    &f_1(\bar{r}) = (2 - \bar{r})\int \dd z\ e^{-z(\bar{r} + 4)} \left(H(z) - \frac{1}{z}\right) + \frac{7}{3}\,,\\
    &f_2(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
    \left(1 + z (2 - \bar{r})\right) \left(H(z) - \frac{1}{z}\right)+ \frac{6}{\bar{r} + 4}\,,\\
    &f_3(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
    \left(2 + z (2 - \bar{r})\right)z \left(H(z) - \frac{1}{z}\right) + \frac{\bar{r} + 10 }{(\bar{r} + 4)^2}\,,\\
\end{align}
where:
\begin{equation}
    H(z) = \sum_{\ell = 1}^{\infty} (2\ell + 3)e^{-z (\ell(\ell + 3) - 4)}\,.
\end{equation}

The functions $f_i$ for the sphere are also displayed in
fig.~\ref{fig:f_functions}. They diverge for $\bar{r} \to -4$, signaling the
instability of the perturbative vacuum, and they go to zero for $\bar{r} \to
\infty$.

@@

\bibitem{WeinbergEffectiveAction}
S. Weinberg,
\textit{The quantum theory of fields, Vol. 2}, Cambridge University Press, (1996) pg. 63.

\bibitem{Brown1980}
Lowell S. Brown, John C. Collins,
\textit{Dimensional renormalization of scalar field theory in curved space-time},
\href{https://doi.org/10.1016/0003-4916(80)90232-8}{Ann. Phys. \textbf{130} (1980) 215}.
\end{thebibliography}

@@
